---
title: "Week 3: Random Variables and Distributions"
subtitle: "PS 812 - Introduction to Statistical Methods in Political Science"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "September 22, 2026"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 812 - University of Wisconsin-Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
code:
  execute: 
    - r
    - python
embed-resources: true
bibliography: ../ps812.bib
link-citations: true

---

## Welcome back! {.title-slide}

## Week overview

::: {.incremental}

- Last week
  - Conditional probability
  - Independence and conditional independence
  - Bayes' rule
- This week
  - **Random variables**
  - **Distributions**
  - **Properties of random variables**

:::

## Where we're going

::: {.incremental}

- This week, we'll go beyond **events** and illustrate a way of talking about random processes that map on to the **real numbers** $\mathbb{R}$
  - These are what we'll call **random variables**
- Random variables are defined by their **distributions**
  - We'll summarize features of these distributions with quantities like the **expectation** and the **variance**
- These will eventually be our empirical **estimands**
  - The share of residents of Wisconsin who turn out in the 2026 midterm election.
  - The share of residents of a given Wisconsin precinct who turn out in the 2026 midterm election **given** that precinct's 2024 Harris vote share.
- We will use our observed data to provide **estimates** of these quantities
  - Assume our observations came from the distribution of interest...how "close" is our estimate from the truth.
  - How can we talk about this when we **don't know** the truth!?

:::

## Random variables

::: {.incremental}

- There are clear limitations in using the language of **events** alone in talking about probabilities
- **Example:** Suppose we're trying to measure **income** using a random **survey**
  - Let $A_{ik}$ be the event that the respondent $i$ has an income of exactly $k$ dollars.
  - How would we represent the probability that a randomly selected respondent has an income between \$50,000 and \$100,000?

    $$
    \begin{aligned}
    P\left(\bigcup_{k = 50{,}000}^{100{,}000} A_{ik}\right) &= \sum_{k = 50{,}000}^{100{,}000} P(A_{ik}) \\
    &= P(A_{i,50000}) + P(A_{i,50001}) + \cdots + P(A_{i,99999}) + P(A_{i,100000})
    \end{aligned}
    $$

- Suppose we measure in **dollars** but want to calculate the probability that the respondent has an income between **£50,000 and £100,000**?
  - Suppose the exchange rate is itself **random**?

- There has to be a better way!

:::

## Random variables

- A **random variable** is a function that maps from the *sample space* $\Omega$ to the *real numbers* $\mathbb{R}$
  - Returning to our income example, we can now say that $X_i$ is a random variable denoting respondent $i$'s income in dollars
  - Now we can write the probability of our earlier event as $P(50{,}000 \le X_i \le 100{,}000)$

  ![](assets/rv_mapping.svg){fig-align="center" width="85%"}


## Random variables

::: {.incremental}

-  We can construct random variables as functions other random variables!
  - Let $R$ be a random variable denoting the GBP-USD exchange rate on a given day.
  - Let $Y_i = R \times X_i$
  - The probability that the respondent has an income between **£50,000 and £100,000** is just 
    $$P (50{,}000 \le Y_i \le 100{,}000) $$

  - $Y_i$ is itself another random variable!

:::

## Random variables: Notation

::: {.incremental}

- We typically refer to **random variables** using capital roman letters
  - (e.g. $X$, $Y$, $D$)
  - Sometimes subscripted when we have repetitions across some margin ($i$ for units or $t$ for time)
- We'll refer to a **realization** of a random variable using lower-case roman letters
  - e.g. $P(X = x)$ refers to the probability that a random variable $X$ takes on a specific value $x$.

:::

## Distribution functions


## Discrete random variables

## Probability mass functions (PMF)


## Bernoulli distribution

## Binomial distribution

## Binomial example: Vote buying in Mexico

- The **2012 Mexico Panel Study** [@greene2016mexico] used a **list experiment** to measure vote buying in the 2012 presidential election
  - Exchanging a vote for a gift is illegal, so respondents may not admit to it when asked directly
- Respondents are asked **how many** items on a list they did in recent weeks (not **which** ones)
  - See a television newscast that mentions a candidate
  - Attend a campaign event
  - Talk about politics with other people
  - *Treated group only:* Exchange your vote for a gift, favor, or access to a service
- Respondents are randomly assigned to the 3-item (**control**) or 4-item (**treated**) list
- Every respondent is also asked the vote-buying question **directly**

## Example: Vote buying in Mexico

```{r}
#| echo: true
#| message: false

library(tidyverse)

# y: number of items; treat: 1 = 4-item list; direct: admitted vote buying when asked directly
mexico <- read_csv("data/mexico_list.csv")
```

## Fitting a Binomial {.app-slide}

::: {.app-frame data-app="binomial_app.html" data-title="Fitting a Binomial to the list experiment"}
:::

::: {.attribution}
2012 Mexico Panel Study, Wave 2 [@greene2016mexico]
:::

```{=html}
<script>
// The widget is a separate page framed here. The <iframe> is built at runtime
// because `embed-resources: true` would otherwise inline the target into a
// data: URI, breaking the app's relative asset paths and its service worker.
// Mounting only once the slide is reached also delays the webR download.
(function () {
  function mount() {
    document.querySelectorAll("section.present .app-frame[data-app]").forEach(function (el) {
      if (el.dataset.mounted) return;
      el.dataset.mounted = "1";
      var f = document.createElement("iframe");
      f.src = el.dataset.app;
      f.title = el.dataset.title || "Interactive widget";
      f.setAttribute("allow", "clipboard-write");
      el.appendChild(f);
    });
  }
  // Reveal is not defined while this script is parsed, so wire up after load.
  window.addEventListener("load", function () {
    if (window.Reveal && typeof Reveal.on === "function") Reveal.on("slidechanged", mount);
    mount();
  });
})();
</script>
```
## Estimating the sensitive-item probability

::: {.incremental}

- The two lists share the same three non-sensitive items, so their means differ by only the sensitive item:

  $$
  \begin{aligned}
  E[Y_i \mid D_i = 1] - E[Y_i \mid D_i = 0] &= (3p + \theta) - 3p \\
  &= \theta
  \end{aligned}
  $$

- So the **difference in means** between the treated and control lists estimates $\theta$

```{r}
#| echo: true

mexico |>
  group_by(treat) |>
  summarise(mean_y = mean(y), n = n())

# theta-hat: the treated list carries one extra (sensitive) item
with(mexico, mean(y[treat == 1]) - mean(y[treat == 0]))

# what the direct question picks up
mean(mexico$direct)
```

- The list experiment recovers **much more** vote buying than asking directly

:::

## Continuous random variables

## Probability density functions (PDFs)

## Normal distribution

## Exponential distribution

## Expectation

## Properties of expectations

## Variance

## Properties of variance

## Next week

::: {.incremental}

- Joint distributions
  - Definining the distribution of $X$ and $Y$ (and $Z$, and...) together!
- Conditional distributions
  - **Slicing** the joint distribution at a particular point
- Conditional expectations and the best linear predictor
- Covariances and correlations

:::

## References {.refs-slide}

::: {#refs}
:::
