---
title: "PS 812: Section 3 Exercise"
author: "YOUR NAME"
date: "9-25-2026"
format:
  html:
    embed-resources: true
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r setup}
#| eval: true
library(tidyverse)

set.seed(812)

senate <- read_csv("data/splitticket_senate_races_2026-09-23.csv")
st_seats <- read_csv("data/splitticket_senate_seat_dist_2026-09-23.csv")
```

# Part 1: Simulation basics

## Task 1: seeds

Run `runif(3)` twice in a row. Then run `set.seed(1)` followed by `runif(3)` twice (setting the seed each time). What's different?

```{r}

```

## Task 2: normal draws

Draw 10,000 values from a normal with mean 50 and SD 10. Check the mean and SD of your draws. Use them to estimate $P(X > 70)$ and compare to the exact answer from `pnorm()`.

```{r}

```

## Task 3: exponential draws

Draw 10,000 values from an exponential with rate 0.5. What should the mean be? Make a histogram.

```{r}

```

## Task 4: a for loop

Flip 10 fair coins and count the heads. Use a `for` loop to repeat this 10,000 times, storing the count each time in a preallocated vector. Estimate $P(\text{7 or more heads})$ and compare to `1 - pbinom(6, 10, 0.5)`.

```{r}

```

## Task 5: the same thing as a function

Write a function `count_heads()` that runs one experiment from Task 4 and returns the count. Redo Task 4 with `map_int()` instead of a loop.

```{r}

```

## Task 6: LLN

Draw 5,000 values from a Uniform(0, 1). Compute the running mean (hint: `cumsum()`) and plot it against the number of draws. Where does it end up?

```{r}

```

## Task 7: CLT

Write a function that returns the mean of `n` draws from a Uniform(0, 1). For `n = 1`, `2`, `5`, and `30`, simulate 5,000 sample means and plot their histograms (`facet_wrap(~ n)`). What happens to the shape and the spread as `n` grows?

```{r}

```

# Part 2: Who wins the Senate?

Data: Split Ticket 2026 Senate model, published by *The Argument* and retrieved 9/23/2026 ([article](https://www.theargumentmag.com/p/split-ticket-2026-midterms-model)).

Columns in `senate`:

- `state`, `state_name`
- `dem_name`: the main non-Republican candidate (an independent in ID, MT, NE, SD)
- `rep_name`: Republican nominee
- `independent`: `TRUE` if `dem_name` is an independent
- `margin`: forecast margin in points (positive = Dem/Ind ahead)
- `dem_prob`: forecast probability (0 to 100) that `dem_name` wins
- `rating`: Safe / Likely / Lean / Tossup

Democrats hold 34 seats that are not up this year and need 51 for control. Count an independent win as a Democratic seat.

```{r}
dem_hold   <- 34
seats_need <- 51
n_races    <- nrow(senate)
```

## Task 8: setup

Add a column `p_dem` with the win probability on a 0 to 1 scale. How many of the 35 races do Democrats need to win? Use `count()` to see how many races fall in each rating.

```{r}

```

## Task 9: expected seats

What is the expected number of Democratic seats? You don't need to simulate this. Does it tell you the probability Democrats win control? Why not?

```{r}

```

## Task 10: simulate one election

Write a function `sim_election(p)` that takes a vector of win probabilities, flips one coin per race, and returns the total number of Democratic seats (including the 34 not up). Hint: `runif(length(p)) < p`. Run it a few times.

```{r}

```

## Task 11: simulate many elections

Simulate 100,000 elections two ways: first with a preallocated `for` loop, then with `map_int()`. Check that the mean number of seats matches Task 9.

```{r}

```

## Task 12: probability of control

Estimate $P(\text{Dem seats} \ge 51)$ and its Monte Carlo standard error. Plot the distribution of seats with `geom_bar()`, filled by who controls the Senate.

```{r}

```

## Task 13: from probabilities to margins

Suppose the actual margin in each race is `margin` plus a $N(0, s^2)$ error, so $P(\text{D wins}) = \Phi(\text{margin}/s)$. Find the $s$ that best fits the published probabilities by regressing `qnorm(p_dem)` on `margin` with no intercept (the slope is $1/s$). Leave out races listed at exactly 0% or 100%. Why?

Plot `p_dem` against `margin` and add the fitted curve with `geom_function()`.

```{r}

```

## Task 14: simulate the margins

Write a function that simulates one election by drawing one normal error per race, adding it to `margin`, and counting the races where the result is above 0. Simulate 100,000 elections and estimate $P(\text{control})$. How does it compare to Task 12? Why isn't it exactly the same?

```{r}

```

## Task 15: independents

Redo Task 12 assuming none of the four independents caucus with Democrats. How much does $P(\text{control})$ change?

```{r}

```

## Bonus: a national polling miss

Polls tend to miss in the same direction everywhere. Split the error into a national piece shared by every race and a state piece:

$$
\varepsilon_s = \text{national} + \text{state}_s, \qquad s^2 = \sigma_{\text{nat}}^2 + \sigma_{\text{state}}^2
$$

Set $\sigma_{\text{nat}} = 5$ and pick $\sigma_{\text{state}}$ so the total stays equal to your $s$ from Task 13. Simulate again.

1. Does each race's individual win probability change?
2. Compare the mean, SD, $P(\text{control})$, $P(\text{54+ seats})$, and $P(\text{47 or fewer})$ to Task 14. Why does $P(\text{control})$ barely move while the tails change a lot?
3. `st_seats` has Split Ticket's own seat distribution (`pct` = percent of their simulations). Compute its mean, SD, and $P(\text{control})$. Which of your models is closer?

```{r}

```
