---
title: "PS 812"
subtitle: "Section 2: Tidyverse/Data Management"
author: "John Hicks"
institute: "Dept. of Political Science, UW-Madison"
date: "9-11-2026"

format:
  revealjs:
    theme: default
    slide-number: true
    transition: none
    code-fold: false
    code-line-numbers: true
    chalkboard: false
    width: 1600
    height: 900
    incremental: true

execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

<style>

.reveal {
  font-family: Arial, sans-serif;
}

/* ---------------------------------
   Headings
   --------------------------------- */

.reveal h1,
.reveal h2,
.reveal h3 {
  color: #8B0000;
}

.reveal h2 {
  background-color: #eeeeee;
  padding: 10px 20px;
}

/* ---------------------------------
   Slide numbers
   --------------------------------- */

.reveal .slide-number {
  color: #666666;
  font-size: 0.7em;
}

/* ---------------------------------
   Code
   --------------------------------- */

.reveal pre {
  font-size: 0.65em;
}

.reveal code {
  font-family: "Fira Code", "Courier New", monospace;
}

/* ---------------------------------
   Cascade / incremental reveals
   --------------------------------- */

.reveal .fragment {
  transition: opacity 0.35s ease-in-out;
}

.reveal .fragment.visible {
  opacity: 1;
}

/* ---------------------------------
   Lists
   --------------------------------- */

.reveal ul li,
.reveal ol li {
  margin-bottom: 0.35em;
}

/* ---------------------------------
   Images
   --------------------------------- */

.reveal img {
  max-height: 70vh;
  object-fit: contain;
}

/* ---------------------------------
   Git workflow
   --------------------------------- */

.git-command {
  font-size: 1.15em;
  margin: 0.5em 0;
}

</style>

# Announcements

## Problem Set 1

1. **Problem Set 1 is due September 22, immediately before class (3:30pm).**

---

# Getting Started

## Today

::: {.fragment}
**1. Tidyverse**

Basic functions
:::

::: {.fragment}
**2. Reading in Data**

See how your `.qmd` becomes the document you submit.
:::

::: {.fragment}
**3. Manipulating Data**

Clone, pull, commit, push.
:::


---

# Tidyverse

## Tidyverse

Library of packages with many useful functions

::: {.fragment}
### 1. dplyr

**Manipulate and summarize data**

- `filter()` — keep rows that meet a condition
- `select()` — choose columns
- `mutate()` — create or modify columns
- `arrange()` — sort rows
- `summarize()` — calculate summary statistics
- `group_by()` — perform operations within groups
:::

---

::: {.fragment}


### 2. ggplot2

**Create visualizations**

- `ggplot()` — start a plot
- `geom_point()` — scatterplots
- `geom_line()` — line graphs
- `geom_bar()` / `geom_col()` — bar graphs
- `geom_histogram()` — distributions
- `facet_wrap()` — create separate panels
- `labs()` — add titles and labels
:::

---

::: {.fragment}
### 3. tidyr

**Reshape and organize data**

- `pivot_longer()` — convert wide data to long format
- `pivot_wider()` — convert long data to wide format
- `separate()` — split one column into multiple columns
- `unite()` — combine multiple columns
- `drop_na()` — remove missing observations
:::

---

::: {.fragment}
### 4. stringr

**Work with text**

- `str_detect()` — find whether text contains a pattern
- `str_replace()` — replace text
- `str_extract()` — extract matching text
- `str_to_lower()` / `str_to_upper()` — change capitalization
- `str_length()` — count characters
:::

---

::: {.fragment}
### 5. readr

**Import and export data**

- `read_csv()` — read a CSV file
- `write_csv()` — save a data frame as a CSV
:::

---

::: {.fragment}
### 6. forcats

**Work with categorical variables**

- `fct_reorder()` — reorder factor levels
- `fct_relevel()` — manually change the order
- `fct_recode()` — rename categories
:::

---

::: {.fragment}
### 7. lubridate

**Work with dates and times**

- `ymd()` — convert text to dates
- `year()` — extract the year
- `month()` — extract the month
- `day()` — extract the day
:::

---

::: {.fragment}
### 8. purrr

**Repeat operations across data**

- `map()` — apply a function repeatedly
- Useful when you need to perform the same operation on multiple objects
:::

---

## The Pipe

The tidyverse (**%>%**) pipe allows you to perform multiple functions sequentially, rather than wrapping them inside each other. 

```{r}
library(tidyverse)
sqrt(2*2)
(2+2) %>% sqrt()

```
You can also use the native pipe without loading tidyverse (**|>**)
```{r}
(2+2) |> sqrt()

```

---

## The Pipe

The true utility of piping comes when you run a complex series of functions. For example:

```{r}
df = mtcars

table(ifelse(scale(log(df$mpg+2))>0,T,F))

```

can instead be written as:
```{r}

df$mpg %>% 
  `+`(2) %>% 
  log() %>% 
  scale() %>% 
  (\(x) ifelse(x > 0, TRUE, FALSE)) %>% 
  table()
```

---

## dplyr Syntax

When you perform functions on objects using the pipe, you automatically select the variable within the object

```{r}

head(df)


df %>%
  filter(mpg > 22) %>% 
  head()


```

---




## Mutate

Mutate can be used to create new variables 

```{r}
# remember to assign the modified df back as an object
df = df %>%
  mutate(log_mpg = log(mpg))

```

## dplyr functions can also be applied not to single variables

```{r}
# Log one variable
df %>%
  mutate(log_mpg = log(mpg)) %>% head()

# Log several variables
df %>%
  mutate(across(c(mpg, wt, hp), log)) %>% head()
```

---

```{r}
# Log all numeric variables
df %>%
  mutate(across(where(is.numeric), log)) %>% head()

# Log variables matching a name pattern
df %>%
  mutate(across(starts_with("x"), log))  %>% head()
```

---

```{r}
# Log variables with an anonymous function
df %>%
  mutate(across(c(mpg, wt), ~ log(.)))  %>% head()

# Log variables and create new names
df %>%
  mutate(across(c(mpg, wt), log, .names = "log_{.col}"))  %>% head()
```

---

```{r}
# Apply multiple functions
df %>%
  mutate(across(c(mpg, wt),
                list(log = log, mean = mean)))  %>% head()

# if_any(): TRUE if any selected variable meets condition
df %>%
  filter(if_any(c(mpg, wt), ~ . > 20))  %>% head()

# if_all(): TRUE if all selected variables meet condition
df %>%
  filter(if_all(c(mpg, wt), ~ . > 20))  %>% head()

```

---

## group_by()/summarise()

Two of the most useful functions in dplyr are group_by() and summarise(). You can use them to compare parts of the data.

Suppose we want to find the average mpg by number of gears. Rather than calculate manually:

```{r}

df %>% 
  group_by(gear) %>% 
  summarise(avg_mpg = mean(mpg,na.rm = T))

```

---

Many functions can be used in summarise, not just mean

```{r}

df %>% 
  group_by(gear) %>% 
  summarise(avg_mpg = sum(mpg,na.rm = T))

```


And we can group by multiple variables

```{r}

df %>% 
  group_by(gear,vs) %>% 
  summarise(avg_mpg = mean(mpg,na.rm = T))

```

---

Note that the resulting object is now "grouped" output. If you want to ungroup it you can use the ungroup() function 

This becomes relevant if you want to group within a seqeunce of functions but not save the full df 

```{r}
## Grouping and Ungrouping

df %>%
  group_by(gear, vs) %>%
  mutate(avg_mpg = mean(mpg, na.rm = TRUE)) %>%
  ungroup() %>% 
  select(mpg,gear,vs,avg_mpg) %>% 
  head()
```

---

## left_join()

`left_join()` is useful when you want to add information from one dataset to another.

Suppose we have a dataset with cars and another dataset with information about the number of cylinders. We can join them using a common variable:

```{r}

df_cyl_fake = df %>% 
  select(cyl) %>% 
  mutate(fake_var = case_when(cyl == 4 ~ "FAKE 4",
                              cyl == 6 ~ "FAKE 6",
                              cyl == 8 ~ "FAKE 8",
                              )) %>% unique()

df %>% 
  left_join(df_cyl_fake, by = "cyl") %>% 
  select(mpg,cyl,fake_var) %>% 
  head()

```

---

## stringr

```{r}

df %>% 
  rownames_to_column() %>% 
  filter(str_detect(rowname, "Toyota"))


```

String syntax is slightly different from normal boolian operators

```{r}
df %>% 
  rownames_to_column() %>% 
  filter(str_detect(rowname, "Toyota|Mazda"))

#draws an error
#df %>% 
#  rownames_to_column() %>% 
 # filter(str_detect(rowname,"Toyota"|"Mazda"))
```

## readr

```{r}
# readr
#df <- read_csv("data.csv") 
#df <- write_csv("data.csv")

#base r
#df <- read.csv("data.csv") 
#df <- write.csv("data.csv")

```


---

## ggplot2

`ggplot2` is useful for **visualizing relationships in your data**.

A basic ggplot follows the structure:

```{r}

df %>% 
  ggplot(aes(x = gear, y = mpg)) +
  geom_point()

```

---

## Common geom_ functions 

1. geom_point() — scatterplot
2. geom_line() — line graph
3. geom_bar() — bar graph
4. geom_histogram() — histogram
5. geom_boxplot() — boxplot
6. labs() - change the labels
7. scale_x_manual - change the x axis

The key idea: aes() tells ggplot what variables to put where; geom_ and other added functions tells it how to display them.

---

## We can combine ggplot with other dplyr functions


```{r}

df %>% 
  group_by(vs,cyl) %>% 
  summarise(avg_mpg = mean(mpg)) %>% 
  mutate(cyl = as.factor(cyl)) %>% 
  ggplot(aes(x=vs,
             y=avg_mpg,
             fill=cyl)) +
  geom_col(position = "dodge")

```

