---
title: "PS 812"
subtitle: "Section 2: Tidyverse/Data Management"
author: "John Hicks"
institute: "Dept. of Political Science, UW-Madison"
date: "9-11-2026"

format:
  revealjs:
    embed-resources: true
    theme: default
    slide-number: true
    transition: none
    code-fold: false
    code-line-numbers: true
    chalkboard: false
    width: 1600
    height: 900
    incremental: true

execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

<style>

.reveal {
  font-family: Arial, sans-serif;
}

/* ---------------------------------
   Headings
   --------------------------------- */

.reveal h1,
.reveal h2,
.reveal h3 {
  color: #8B0000;
}

.reveal h2 {
  background-color: #eeeeee;
  padding: 10px 20px;
}

/* ---------------------------------
   Slide numbers
   --------------------------------- */

.reveal .slide-number {
  color: #666666;
  font-size: 0.7em;
}

/* ---------------------------------
   Code
   --------------------------------- */

.reveal pre {
  font-size: 0.65em;
}

.reveal code {
  font-family: "Fira Code", "Courier New", monospace;
}

/* ---------------------------------
   Cascade / incremental reveals
   --------------------------------- */

.reveal .fragment {
  transition: opacity 0.35s ease-in-out;
}

.reveal .fragment.visible {
  opacity: 1;
}

/* ---------------------------------
   Lists
   --------------------------------- */

.reveal ul li,
.reveal ol li {
  margin-bottom: 0.35em;
}

/* ---------------------------------
   Images
   --------------------------------- */

.reveal img {
  max-height: 70vh;
  object-fit: contain;
}

/* ---------------------------------
   Git workflow
   --------------------------------- */

.git-command {
  font-size: 1.15em;
  margin: 0.5em 0;
}

</style>

# Announcements

## Problem Set 1

1. **Problem Set 1 is due September 22 at 11:59pm.**

---

# Getting Started

## Today

::: {.fragment}
**1. Tidyverse**

The pipe, `filter()`, `select()`, `mutate()`, and `across()`
:::

::: {.fragment}
**2. The CES Data**

Download the Cooperative Election Study and read it in with `haven`.
:::

::: {.fragment}
**3. Manipulating Data**

Clone, pull, commit, push.
:::


---

# Tidyverse

## Tidyverse

Library of packages with many useful functions

::: {.fragment}
### 1. dplyr

**Manipulate and summarize data**

- `filter()` — keep rows that meet a condition
- `select()` — choose columns
- `mutate()` — create or modify columns
- `arrange()` — sort rows
- `summarize()` — calculate summary statistics
- `group_by()` — perform operations within groups
:::

---

::: {.fragment}


### 2. ggplot2 *(later)*

**Create visualizations**

- `ggplot()` — start a plot
- `geom_point()` — scatterplots
- `geom_line()` — line graphs
- `geom_bar()` / `geom_col()` — bar graphs
- `geom_histogram()` — distributions
- `facet_wrap()` — create separate panels
- `labs()` — add titles and labels
:::

---

::: {.fragment}
### 3. tidyr

**Reshape and organize data**

- `pivot_longer()` — convert wide data to long format
- `pivot_wider()` — convert long data to wide format
- `separate()` — split one column into multiple columns
- `unite()` — combine multiple columns
- `drop_na()` — remove missing observations
:::

---

::: {.fragment}
### 4. stringr

**Work with text**

- `str_detect()` — find whether text contains a pattern
- `str_subset()` — keep only the matching elements of a vector
- `str_replace()` — replace text
- `str_to_lower()` / `str_to_upper()` — change capitalization
- `str_length()` — count characters
:::

---

::: {.fragment}
### 5. readr

**Import and export data**

- `read_csv()` — read a CSV file
- `write_csv()` — save a data frame as a CSV
:::

---

::: {.fragment}
### 6. forcats

**Work with categorical variables**

- `fct_reorder()` — reorder factor levels
- `fct_relevel()` — manually change the order
- `fct_recode()` — rename categories
:::

---

::: {.fragment}
### 7. lubridate

**Work with dates and times**

- `ymd()` — convert text to dates
- `year()` — extract the year
- `month()` — extract the month
- `day()` — extract the day
:::

---

::: {.fragment}
### 8. purrr *(later)*

**Repeat operations across data**

- `map()` — apply a function repeatedly
- Useful when you need to perform the same operation on multiple objects
:::

---

## The Pipe

Tidyverse-style code is built around **piping**: passing the result of one function straight into the next, instead of nesting functions inside each other.

The base R **native pipe** (`|>`) ships with R itself:

```{r}
library(tidyverse)
sqrt(2*2)
(2+2) |> sqrt()
```

You will also see the **magrittr pipe** (`%>%`) in tidyverse. It works the same:

```{r}
(2+2) %>% sqrt()
```

---

## The value of piping

Piping becomes valuable when you chain several steps together. Compare:

```{r}
table(ifelse(scale(log(mtcars$mpg + 2)) > 0, T, F))
```

The same computation, piped step by step with the native pipe:

```{r}
mtcars$mpg |>
  (\(x) x + 2)() |>
  log() |>
  scale() |>
  (\(x) ifelse(x > 0, TRUE, FALSE))() |>
  table()
```

---

# The CES Data

## Downloading the CES

::: {.fragment}
**1.** Go to the CES Dataverse page:
[dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/DFUGJR](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/DFUGJR)
:::

::: {.fragment}
**2.** Download **`CES25_Common.dta`** — a Stata file, not a `.csv`
:::

::: {.fragment}
**3.** Save it into your project folder, next to your `.qmd` file
:::

---

## The `haven` package

PS data is often distributed as Stata (`.dta`), SPSS (`.sav`), or SAS files rather than `.csv`, which base R can't read without packages like `haven`

```{r}
#| eval: false
library(haven)

ces_full <- read_dta("CES25_Common.dta")

head(ces_full)
```

---

## `haven` and labelled data

Stata variables often carry value labels (e.g. `1 = "Democrat"`). `haven` preserves these as a special `labelled` class instead of silently turning them into plain numbers or text.

```{r}
#| eval: false
class(ces_full$pid3)
attributes(ces_full$pid3)

# turn the value labels into a factor
as_factor(ces_full$pid3) |> head()
```

`as_factor()` uses the stored labels; `zap_labels()` throws the labels away and keeps the raw numeric codes.

---

## Reading in what we'll use today

```{r}
library(haven)

ces <- read_dta("data/CES25_Common.dta")

head(ces)
```

---

## dplyr Syntax

When you pipe a data frame into a dplyr function, you call variables directly without using the selector `$`

```{r}
ces |> select(pid3, commonweight, CC25_321a) |> head()

ces |>
  filter(pid3 == 1) |>
  select(pid3, commonweight, CC25_321a) |>
  head()
```

---

## Subsetting: filter() and select()

`filter()` keeps **rows**; `select()` keeps **columns**. Selecting is useful for preserving memory

```{r}
# rows: independents only
ces |>
  filter(pid3 == 3) |>
  select(pid3, starts_with("CC25_321")) |>
  head()

# columns: drop everything from the "34" block of items
ces |>
  select(-starts_with("CC25_34")) |>
  names()
```

---

## Mean of an indicator

Every policy item here is already coded as an **indicator**: 1 = support, 0 = oppose, `NA` = no answer. The mean of a 0/1 variable is the proportion of 1s or the share who support.

```{r}
# share of respondents who support banning assault rifles
mean(ces$CC25_321a == 1, na.rm = TRUE)

# same thing, dplyr style
ces |>
  summarise(support = mean(CC25_321a == 1, na.rm = TRUE))
```

---

## mutate()

`mutate()` builds new variables. We can use this to create an indicator out of any logical condition. Remember to reassignthe new dataframe to have it stored.

```{r}
ces = ces |>
  mutate(dem = pid3 == 1)

# or 

ces = ces |>
  mutate(dem2 = if_else(pid3 == 1,1,0))


mean(ces$dem, na.rm = TRUE)
mean(ces$dem2, na.rm = TRUE)

ces |>
  summarise(share_dem = mean(dem, na.rm = TRUE))

```

---

## across() — apply one function to many columns

`across()` runs the same function over several columns inside `mutate()` or `summarise()`, instead of retyping it once per column.

```{r}
# share supporting each gun control item
ces |>
  summarise(across(starts_with("CC25_321"), \(x) mean(x == 1, na.rm = TRUE)))

# same idea, but for every numeric column
ces |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE))) |>
  select(1:5)
```

---

```{r}
# custom names for each result
ces |>
  summarise(across(starts_with("CC25_326"),
                    \(x) mean(x == 1, na.rm = TRUE),
                    .names = "support_{.col}"))

# across() inside mutate() instead of summarise()
ces |>
  mutate(across(starts_with("CC25_343"), ~ .x == 1, .names = "{.col}_yes")) |>
  select(ends_with("_yes")) |>
  head()
```

---

```{r}
# if_any(): TRUE if any selected variable meets the condition
ces |>
  filter(if_any(starts_with("CC25_342"), ~ . == 1)) |>
  nrow()

# if_all(): TRUE only if every selected variable meets the condition
ces |>
  filter(if_all(starts_with("CC25_342"), ~ . == 1)) |>
  nrow()
```

---

## group_by()/summarise()

Two of the most useful functions in dplyr are group_by() and summarise(). You can use them to compare parts of the data. To demo this we'll switch to a simpler dataset, `mtcars`

Suppose we want to find the average mpg by number of gears. Rather than calculate manually:

```{r}
df = mtcars

df %>%
  group_by(gear) %>%
  summarise(avg_mpg = mean(mpg,na.rm = T))

```

---

Many functions can be used in summarise, not just mean

```{r}

df %>%
  group_by(gear) %>%
  summarise(avg_mpg = sum(mpg,na.rm = T))

```


And we can group by multiple variables

```{r}

df %>%
  group_by(gear,vs) %>%
  summarise(avg_mpg = mean(mpg,na.rm = T))

```

---

Note that the resulting object is now "grouped" output. If you want to ungroup it you can use the ungroup() function

This becomes relevant if you want to group within a seqeunce of functions but not save the full df

```{r}
## Grouping and Ungrouping

df %>%
  group_by(gear, vs) %>%
  mutate(avg_mpg = mean(mpg, na.rm = TRUE)) %>%
  ungroup() %>%
  select(mpg,gear,vs,avg_mpg) %>%
  head()
```

---

## left_join()

`left_join()` is one of many functions that adds columns from one data frame onto another using some shared variable(s) as a key

```{r}
party_labels = tibble(
  pid3 = c(1, 2, 3),
  party_name = c("Democrat", "Republican", "Independent")
)

ces |>
  left_join(party_labels, by = "pid3") |>
  select(pid3, party_name, commonweight) |>
  head()
```

---

## stringr

With 300+ variables in the full CES, you can use stringr to search for variables.

```{r}
# every gun control item
names(ces) |> str_subset("CC25_321")

# every column whose name contains "34"
names(ces) |> str_subset("34")

# regex alternation with |: gun control OR environment items
names(ces) |> str_subset("CC25_321|CC25_326")
```

