---
title: "PS 812: Section 2 Exercise"
author: "YOUR NAME"
date: "9-18-2026"
format:
  html:
    embed-resources: true
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r setup}
#| eval: true
library(tidyverse)
library(haven)


```

## Task 1: Read and inspect

Read in `CES25_Common.dta` with `haven` (already done above). Use `glimpse()` or `names()` to look at what's in it.

```{r}

```

## Task 2: labelled data

Check `class()` and `attributes()` on `pid3`. Convert it to a factor with `as_factor()`.

```{r}

```

## Task 3: filter()

Subset `ces` to Republicans only. How do you know which value represents Republicans?

```{r}


```

## Task 4: select()

Select `pid3`, `commonweight`, and every gun control item (`CC25_321a`–`CC25_321e`).

```{r}

```

## Task 5: filter() + select() together

Subset to Independents and select just the immigration items (`CC25_323a`–`CC25_323d`).

```{r}

```

## Task 6: mean of an indicator

Compute the share of respondents who support `CC25_323c` (build the border wall). Remember it's coded 1 = Support, 2 = Oppose.

```{r}

```

## Task 7: mutate()

Create a new column `supports_wall` that is `TRUE` when `CC25_323c == 1`. Then create another variable `supports_wall_cat` with categorical values instead of numeric. You can pick the values.

```{r}

```

## Task 8: across()

Use `across()` inside `summarise()` to compute the share supporting every "Trump 2nd term" item (`CC25_342a`–`CC25_342i`) at once.

```{r}

```

## Task 9: group_by() + summarise()

Compute the share supporting `CC25_326d` (increase fossil fuel production), broken out by `pid3`.

```{r}

```

## Task 10: if_any() / if_all()

How many respondents support **at least one** of the "Health & housing" items (`CC25_328a`–`CC25_328e`)? How many support **all** of them?

```{r}

```

## Task 11: left_join()

Build a small `tibble` that maps `pid3` codes (1, 2, 3) to labels ("Democrat", "Republican", "Independent"), then join it onto `ces`.

```{r}

```

## Task 12: stringr

Use `str_subset()` on `names(ces)` to find every variable belonging to the "Congress" block (`CC25_340*`).

```{r}

```

## Task 13: poke around

Pick one variable in `ces` we haven't used yet. Figure out what it means (`attributes()`, value labels) and compute something with it.

```{r}

```
