---
title: "Week 4: Joint and Conditional Distributions"
subtitle: "PS 812 - Introduction to Statistical Methods in Political Science"
author:
  - name: "Anton Strezhnev"
    affiliation: "University of Wisconsin-Madison"
    email: strezhnev@wisc.edu
date: "September 29, 2026"
date-format: "MMMM D, YYYY"
editor: 
  render-on-save: true
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    overview: true
    navigation-mode: vertical
    progress: false
    controls: true
    link-external-newwindow: true
    footer: "PS 812 - University of Wisconsin-Madison"
    html-math-method:
      method: mathjax
      url: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js
execute: 
  cache: false
code:
  execute: 
    - r
    - python
embed-resources: true
bibliography: ../ps812.bib
link-citations: true

---

## Welcome back! {.title-slide}

<!-- MathJax macros for the whole deck. The reveal math plugin overwrites
     window.MathJax, so a header config would be clobbered; defining them in
     the first slide registers them before any later slide is typeset. The
     wrapper takes up no space and renders nothing. -->
::: {style="position:absolute;width:0;height:0;overflow:hidden;opacity:0"}
$$
\newcommand{\Var}{\text{Var}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\Corr}{\text{Corr}}
\newcommand{\indep}{\mathrel{\perp\mkern-10mu\perp}}
$$
:::

## Week overview

::: {.incremental}

- Last week
  - Random variables
  - Distributions: CDFs, PMFs and PDFs
  - Expectations and variances
- This week
  - **Joint distributions**
  - **Conditional distributions**
  - **Conditional expectations**
  - **Covariances and correlations**

:::

## From one random variable... {.bvn-slide}

```{r}
#| echo: false

library(plotly)

# The 2D warm-up, one click per row: the mean moves up, down, and back to 0,
# then the SD widens, narrows, and returns to 1
bvn_walk <- list(c(mu = 0, sd = 1),
                 c(mu = 1, sd = 1), c(mu = -1, sd = 1), c(mu = 0, sd = 1),
                 c(mu = 0, sd = 1.5), c(mu = 0, sd = 0.7), c(mu = 0, sd = 1))

# Correlations the joint density steps through, one per click once it's up
bvn_rhos <- c(0, 0.5, 0.9, -0.5, -0.9)

# Top of the density axis: the 2D stages need room for the SD = 0.7 peak (0.57)
ztop_2d <- 0.6
ztop_3d <- 0.42

lim <- 3   # x and y run over [-lim, lim]
g <- seq(-lim, lim, length.out = 61)
n <- length(g)
phi <- dnorm(g)
wall <- lim - 0.02   # just in front of the axis walls, so the curves don't z-fight them

# Shaded area under a marginal, as a flat strip of triangles between the floor
# (vertices 1..n) and the curve (vertices n+1..2n)
strip <- list(i = c(0:(n - 2), 1:(n - 1)),
              j = c(1:(n - 1), (n + 1):(2 * n - 1)),
              k = c(n:(2 * n - 2), n:(2 * n - 2)))
n_tri <- length(strip$i)
flat <- list(ambient = 1, diffuse = 0, specular = 0, fresnel = 0)

bvn_axis <- function(title, range, ticks, ...) {
  list(title = list(text = title), range = range, tickvals = ticks,
       gridcolor = "#d9d9d9", zeroline = FALSE, showspikes = FALSE,
       showbackground = TRUE, backgroundcolor = "#EFEFEF",
       tickfont = list(size = 12), tickangle = 0, ...)
}

bvn_label <- function(x, y, text, colour, visible = TRUE) {
  list(x = x, y = y, z = 0.3, text = text, showarrow = FALSE, xanchor = "left",
       font = list(size = 16, color = colour), visible = visible)
}

# Trace order matters: bivariate_normal.js addresses traces by index
p_bvn <- plot_ly(width = 1000, height = 560) |>
  add_surface(x = g, y = g, z = matrix(0, n, n), visible = FALSE,
              cmin = 0, cmax = 0.37, showscale = FALSE, hoverinfo = "skip",
              # purple: the red of X mixed with the blue of Y
              colorscale = list(c(0, "#e9dff1"), c(0.35, "#9b6cc2"), c(1, "#4a1766")),
              lighting = list(ambient = 0.7, diffuse = 0.5, specular = 0.1,
                              roughness = 0.8, fresnel = 0.05),
              contours = list(x = list(highlight = FALSE), y = list(highlight = FALSE),
                              z = list(show = TRUE, start = 0.02, end = 0.4, size = 0.04,
                                       color = "#ffffff", width = 1, highlight = FALSE))) |>
  # X's marginal on the back wall
  add_trace(type = "mesh3d", x = c(g, g), y = rep(wall, 2 * n), z = c(rep(0, n), phi),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#c5050c", n_tri), opacity = 0.15,
            flatshading = TRUE, lighting = flat, hoverinfo = "skip") |>
  add_trace(type = "scatter3d", mode = "lines", x = g, y = rep(wall, n), z = phi,
            line = list(color = "#c5050c", width = 7), hoverinfo = "skip") |>
  # Y's marginal on the side wall, grown by the animation
  add_trace(type = "mesh3d", x = rep(-wall, 2 * n), y = c(g, g), z = rep(0, 2 * n),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#0479a8", n_tri), opacity = 0.15,
            flatshading = TRUE, lighting = flat, hoverinfo = "skip", visible = FALSE) |>
  add_trace(type = "scatter3d", mode = "lines", x = rep(-wall, n), y = g, z = rep(0, n),
            line = list(color = "#0479a8", width = 7), hoverinfo = "skip", visible = FALSE) |>
  layout(
    showlegend = FALSE,
    paper_bgcolor = "#F7F7F7",
    margin = list(l = 0, r = 0, t = 0, b = 0),
    font = list(family = "Red Hat Text, sans-serif", color = "#333333"),
    scene = list(
      xaxis = bvn_axis("x", c(-lim, lim), -lim:lim),
      yaxis = bvn_axis("", c(-lim, lim), -lim:lim, showticklabels = FALSE),
      zaxis = bvn_axis("", c(0, ztop_2d), seq(0.1, 0.6, 0.1)),
      aspectmode = "manual", aspectratio = list(x = 1, y = 1, z = 0.7),
      camera = list(projection = list(type = "orthographic")),   # set by the animation
      dragmode = "turntable", hovermode = FALSE,
      annotations = list(
        bvn_label(1.2, wall, "f<sub>X</sub>(x)", "#c5050c"),
        bvn_label(-wall, 1.2, "f<sub>Y</sub>(y)", "#0479a8", visible = FALSE)
      )
    )
  ) |>
  config(displayModeBar = FALSE, doubleClick = FALSE)

bvn_base <- p_bvn   # the bare widget, reused by the correlation reprise

p_bvn <- htmlwidgets::onRender(
  p_bvn,
  paste(readLines("bvn-anim/bivariate_normal.js"), collapse = "\n"),
  data = list(grid = g, rhos = bvn_rhos, walk = lapply(bvn_walk, as.list),
              ztop2d = ztop_2d, ztop3d = ztop_3d)
)

# One empty fragment per animation step after the first: the rest of the
# walk, the swing into 3D, and one per correlation
htmltools::tagList(
  p_bvn,
  lapply(seq_len(length(bvn_walk) + length(bvn_rhos)),
         function(i) htmltools::span(class = "fragment bvn-step"))
)
```

## Joint distribution functions

::: {.incremental}

- We've so far talked about the joint probabilities of **events** $P(A \cap B)$
  - And their **conditional** probabilities: $P(A \mid B)$

- Let's extend these ideas to talk about **random variables**
- We can define a **joint cumulative distribution function** that takes values of *multiple* variables as inputs

  - For two random variables, $X$ and $Y$

    $$F_{XY}(x, y) = P(X \le x, Y \le y) $$

  - We can generalize to **many** random variables

    $$F_{X_1, \dotsc X_n}(x_1, \dotsc x_n) = P(X_1 \le x_1, \dotsc X_n \le x_n)$$

:::

## A joint CDF {.cdf-slide}

```{r}
#| echo: false

# The CDF of the opening animation's rho = 0.5 bivariate normal, in the same
# style, with the grid, strips and axis helpers from that chunk. The marginal
# CDFs sit faint on the back walls: F(x, y) climbs to F_X(x) as y grows, and
# to F_Y(y) as x grows.
rho_cdf <- 0.5
F_cdf <- outer(g, g, Vectorize(function(y, x)
  mvtnorm::pmvnorm(upper = c(x, y), corr = matrix(c(1, rho_cdf, rho_cdf, 1), 2),
                   algorithm = mvtnorm::TVPACK())[1]))   # rows are y, as plotly wants
Phi <- pnorm(g)

# The opening animation's resting camera: azimuth -62, elevation 34, zoom 1.2
cdf_az <- -62 * pi / 180
cdf_el <- 34 * pi / 180
cdf_zoom <- 1.2

cdf_quiet <- list(width = 3.5, line = 0.45, fill = 0.06, label = 0.55)

plot_ly(width = 1000, height = 560) |>
  add_surface(x = g, y = g, z = F_cdf, cmin = 0, cmax = 1,
              showscale = FALSE, hoverinfo = "skip",
              colorscale = list(c(0, "#e9dff1"), c(0.35, "#9b6cc2"), c(1, "#4a1766")),
              lighting = list(ambient = 0.7, diffuse = 0.5, specular = 0.1,
                              roughness = 0.8, fresnel = 0.05),
              contours = list(x = list(highlight = FALSE), y = list(highlight = FALSE),
                              z = list(show = TRUE, start = 0.1, end = 0.9, size = 0.1,
                                       color = "#ffffff", width = 1, highlight = FALSE))) |>
  add_trace(type = "mesh3d", x = c(g, g), y = rep(wall, 2 * n), z = c(rep(0, n), Phi),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#c5050c", n_tri),
            opacity = cdf_quiet$fill, flatshading = TRUE, lighting = flat, hoverinfo = "skip") |>
  add_trace(type = "scatter3d", mode = "lines", x = g, y = rep(wall, n), z = Phi,
            opacity = cdf_quiet$line, line = list(color = "#c5050c", width = cdf_quiet$width),
            hoverinfo = "skip") |>
  add_trace(type = "mesh3d", x = rep(-wall, 2 * n), y = c(g, g), z = c(rep(0, n), Phi),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#0479a8", n_tri),
            opacity = cdf_quiet$fill, flatshading = TRUE, lighting = flat, hoverinfo = "skip") |>
  add_trace(type = "scatter3d", mode = "lines", x = rep(-wall, n), y = g, z = Phi,
            opacity = cdf_quiet$line, line = list(color = "#0479a8", width = cdf_quiet$width),
            hoverinfo = "skip") |>
  layout(
    showlegend = FALSE,
    paper_bgcolor = "#F7F7F7",
    margin = list(l = 0, r = 0, t = 0, b = 0),
    font = list(family = "Red Hat Text, sans-serif", color = "#333333"),
    scene = list(
      xaxis = bvn_axis("x", c(-lim, lim), -lim:lim),
      yaxis = bvn_axis("y", c(-lim, lim), -lim:lim),
      zaxis = bvn_axis("", c(0, 1.05), seq(0.25, 1, 0.25)),
      aspectmode = "manual",
      aspectratio = list(x = cdf_zoom, y = cdf_zoom, z = 0.7 * cdf_zoom),
      camera = list(eye = list(x = 1.25 * cos(cdf_el) * cos(cdf_az),
                               y = 1.25 * cos(cdf_el) * sin(cdf_az),
                               z = 1.25 * sin(cdf_el)),
                    projection = list(type = "orthographic")),
      dragmode = "turntable", hovermode = FALSE,
      annotations = list(
        c(list(x = 0.6, y = wall, z = 0.86, text = "F<sub>X</sub>(x)", showarrow = FALSE,
               xanchor = "left", font = list(size = 16, color = "#c5050c")),
          opacity = cdf_quiet$label),
        c(list(x = -wall, y = 0.6, z = 0.86, text = "F<sub>Y</sub>(y)", showarrow = FALSE,
               xanchor = "left", font = list(size = 16, color = "#0479a8")),
          opacity = cdf_quiet$label)
      )
    )
  ) |>
  config(displayModeBar = FALSE, doubleClick = FALSE)
```

## Joint PMFs

::: {.incremental}

- For discrete random variables, we can also talk about the **joint** probability that $X$ takes on a particular value $x$ and $Y$ takes on a particular value $y$

- We define the **joint** probability mass function (PMF) in a similar way as before

  - For two discrete random variables, $X$ and $Y$

    $$p_{XY}(x, y) = P(X = x, Y = y) $$

  - Generalizing to **many** random variables

    $$p_{X_1, \dotsc X_n}(x_1, \dotsc x_n) = P(X_1 = x_1, \dotsc X_n = x_n) $$
:::

## A discrete joint distribution {.pmf-slide}

```{r}
#| echo: false

# The week 3 list experiment: D is the list assignment (1 = treated), Y the
# number of items the respondent reported
mexico <- read.csv("../week3/data/mexico_list.csv")
pmf_d <- mexico$treat
pmf_y <- mexico$y
pmf_n <- nrow(mexico)

# Every respondent is one dot, stacked in a column over their (d, y) cell, so
# a column's height is the share of respondents in that cell: the empirical
# joint PMF P(D = d, Y = y). Each layer of a column is a 3 x 3 block of dots.
per_side <- 3
per_layer <- per_side^2
layer_h <- per_layer / pmf_n   # one layer of dots, in probability units

# The x and y axes cover ranges of different lengths, so a square footprint
# needs different spacings in data units. pmf_aspect is the scene's shape.
pmf_xr <- c(-0.5, 1.5)
pmf_yr <- c(-0.5, 4.5)
pmf_aspect <- list(x = 1, y = 1.8, z = 1.1)
foot <- 0.026   # dot spacing, in the scene's own units (the ones aspectratio uses)
dx <- foot * diff(pmf_xr) / pmf_aspect$x
dy <- foot * diff(pmf_yr) / pmf_aspect$y

cell <- paste(pmf_d, pmf_y)
k <- ave(seq_len(pmf_n), cell, FUN = seq_along) - 1   # position within the column
pos <- k %% per_layer
dots <- data.frame(
  d = pmf_d,
  x3 = pmf_d + (pos %% per_side - (per_side - 1) / 2) * dx,
  y3 = pmf_y + (pos %/% per_side - (per_side - 1) / 2) * dy,
  z3 = (k %/% per_layer + 0.5) * layer_h
)

# The PMF itself, for the labels over each column. The control list has only
# three items, so (0, 4) is an impossible cell with probability 0.
pmf <- expand.grid(d = 0:1, y = 0:4)
pmf$p <- mapply(function(d, y) mean(pmf_d == d & pmf_y == y), pmf$d, pmf$y)

pmf_axis <- function(title, ...) {
  list(title = list(text = title), gridcolor = "#d9d9d9", zeroline = FALSE,
       showspikes = FALSE, showbackground = TRUE, backgroundcolor = "#EFEFEF",
       tickfont = list(size = 12), tickangle = 0, ...)
}

pmf_labels <- lapply(seq_len(nrow(pmf)), function(i) {
  list(x = pmf$d[i], y = pmf$y[i], z = pmf$p[i] + 0.012,
       text = sprintf("%.3f", pmf$p[i]), showarrow = FALSE, yanchor = "bottom",
       font = list(size = 13, color = "#333333"))
})

pmf_cols <- c("#0479A8", "#c5050c")   # control blue, treated red, as in week 3

p_pmf <- plot_ly(width = 1000, height = 560)
for (d in 0:1) {
  s <- dots[dots$d == d, ]
  p_pmf <- p_pmf |>
    add_trace(type = "scatter3d", mode = "markers", x = s$x3, y = s$y3, z = s$z3,
              marker = list(size = 6, color = pmf_cols[d + 1], opacity = 1,
                            line = list(color = "#ffffff", width = 0.5)),
              hoverinfo = "skip")
}

p_pmf |>
  layout(
    showlegend = FALSE,
    paper_bgcolor = "#F7F7F7",
    margin = list(l = 0, r = 0, t = 0, b = 0),
    font = list(family = "Red Hat Text, sans-serif", color = "#333333"),
    scene = list(
      xaxis = pmf_axis("D (list assignment)", range = pmf_xr, tickvals = 0:1,
                       ticktext = c("0 control", "1 treated")),
      yaxis = pmf_axis("Y (items reported)", range = pmf_yr, tickvals = 0:4),
      zaxis = pmf_axis("P(D = d, Y = y)", range = c(0, 0.27),
                       tickvals = seq(0, 0.25, 0.05)),
      aspectmode = "manual", aspectratio = pmf_aspect,
      # from about 45 degrees, so the control row shows between the treated columns
      camera = list(eye = list(x = 0.95, y = -1.05, z = 0.75),
                    projection = list(type = "orthographic")),
      dragmode = "turntable", hovermode = FALSE,
      annotations = pmf_labels
    )
  ) |>
  config(displayModeBar = FALSE, doubleClick = FALSE)
```

## Joint PDFs

::: {.incremental}

- As before, with continuous random variables $X$ and $Y$, the probability that $X$ or $Y$ take on specific values $x$ and $y$ is zero.
  - We cannot work with a probability mass function $P(X = x, Y = y)$

- Instead, we'll define a **joint density function** as the derivative of the joint CDF with respect to each of the variables

  - For two continuous random variables, $X$ and $Y$

    $$f_{XY}(x, y) = \frac{\partial^2 F_{XY}(x,y)}{\partial x \partial y} $$

  - Again, generalizing to **many** random variables

    $$f_{X_1, \dotsc X_n}(x_1, \dotsc x_n) = \frac{\partial^n F_{X_1, \dotsc X_n}(x_1, \dotsc x_n)}{\partial x_1 \dotsc \partial x_n} $$

:::

## A continuous joint distribution {.pdf-slide}

```{r}
#| echo: false

# The opening animation's rho = 0.5 stage, held still so it can be rotated.
# Reuses the grid, strips and axis helpers from that chunk, and matches its
# final look: the surface in front, the marginals faint on the back walls.
rho_pdf <- 0.5
f_pdf <- outer(g, g, function(y, x)
  exp(-(x^2 - 2 * rho_pdf * x * y + y^2) / (2 * (1 - rho_pdf^2))) /
    (2 * pi * sqrt(1 - rho_pdf^2)))   # rows are y, as plotly wants
peak_pdf <- max(f_pdf)

# The animation's resting camera: azimuth -62, elevation 34, zoom 1.2
cam_az <- -62 * pi / 180
cam_el <- 34 * pi / 180
cam_zoom <- 1.2

quiet <- list(width = 3.5, line = 0.45, fill = 0.06, label = 0.55)

plot_ly(width = 1000, height = 560) |>
  add_surface(x = g, y = g, z = f_pdf, cmin = 0, cmax = peak_pdf,
              showscale = FALSE, hoverinfo = "skip",
              colorscale = list(c(0, "#e9dff1"), c(0.35, "#9b6cc2"), c(1, "#4a1766")),
              lighting = list(ambient = 0.7, diffuse = 0.5, specular = 0.1,
                              roughness = 0.8, fresnel = 0.05),
              contours = list(x = list(highlight = FALSE), y = list(highlight = FALSE),
                              z = list(show = TRUE, start = 0.02, end = 0.4, size = 0.04,
                                       color = "#ffffff", width = 1, highlight = FALSE))) |>
  add_trace(type = "mesh3d", x = c(g, g), y = rep(wall, 2 * n), z = c(rep(0, n), phi),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#c5050c", n_tri),
            opacity = quiet$fill, flatshading = TRUE, lighting = flat, hoverinfo = "skip") |>
  add_trace(type = "scatter3d", mode = "lines", x = g, y = rep(wall, n), z = phi,
            opacity = quiet$line, line = list(color = "#c5050c", width = quiet$width),
            hoverinfo = "skip") |>
  add_trace(type = "mesh3d", x = rep(-wall, 2 * n), y = c(g, g), z = c(rep(0, n), phi),
            i = strip$i, j = strip$j, k = strip$k, facecolor = rep("#0479a8", n_tri),
            opacity = quiet$fill, flatshading = TRUE, lighting = flat, hoverinfo = "skip") |>
  add_trace(type = "scatter3d", mode = "lines", x = rep(-wall, n), y = g, z = phi,
            opacity = quiet$line, line = list(color = "#0479a8", width = quiet$width),
            hoverinfo = "skip") |>
  layout(
    showlegend = FALSE,
    paper_bgcolor = "#F7F7F7",
    margin = list(l = 0, r = 0, t = 0, b = 0),
    font = list(family = "Red Hat Text, sans-serif", color = "#333333"),
    scene = list(
      xaxis = bvn_axis("x", c(-lim, lim), -lim:lim),
      yaxis = bvn_axis("y", c(-lim, lim), -lim:lim),
      zaxis = bvn_axis("", c(0, ztop_3d), seq(0.1, 0.4, 0.1)),
      aspectmode = "manual",
      aspectratio = list(x = cam_zoom, y = cam_zoom, z = 0.7 * cam_zoom),
      camera = list(eye = list(x = 1.25 * cos(cam_el) * cos(cam_az),
                               y = 1.25 * cos(cam_el) * sin(cam_az),
                               z = 1.25 * sin(cam_el)),
                    projection = list(type = "orthographic")),
      dragmode = "turntable", hovermode = FALSE,
      annotations = list(
        c(bvn_label(1.25, wall, "f<sub>X</sub>(x)", "#c5050c"), opacity = quiet$label),
        c(bvn_label(-wall, 1.2, "f<sub>Y</sub>(y)", "#0479a8"), opacity = quiet$label)
      )
    )
  ) |>
  config(displayModeBar = FALSE, doubleClick = FALSE)
```

## Covariance and correlation

::: {.incremental}

- When we have multiple random variables, we may want a **summary** of how the two quantities vary *together*.
  - In the same way we wanted a summary of *central tendency* (**expectation**) or *variability* (**variance**)
- The **covariance** between two random variables $X$ and $Y$ is defined as

  $$ \Cov(X, Y) = \mathbb{E}[(X - \mathbb{E}[X])(Y - \mathbb{E}[Y])] $$

- We can factor and apply linearity of expectations to get another way of writing the covariance

  $$
  \begin{aligned}
  \Cov(X, Y) &= \mathbb{E}[(X - \mathbb{E}[X])(Y - \mathbb{E}[Y])] \\
  &= \mathbb{E}\big[XY - X\,\mathbb{E}[Y] - \mathbb{E}[X]\,Y + \mathbb{E}[X]\,\mathbb{E}[Y]\big] \\
  &= \mathbb{E}[XY] - \mathbb{E}[X]\,\mathbb{E}[Y] - \mathbb{E}[X]\,\mathbb{E}[Y] + \mathbb{E}[X]\,\mathbb{E}[Y] \\
  &= \mathbb{E}[XY] - \mathbb{E}[X]\,\mathbb{E}[Y]
  \end{aligned}
  $$

:::

## Properties of covariance

::: {.incremental}

- We can pull out constants out of the covariance. Let $a$ and $b$ be constants and $X$ and $Y$ random variables:

  $$ \Cov(aX, bY) = ab\Cov(X, Y) $$

- Adding a constant doesn't change the covariance (just as it doesn't change the variance)

  $$ \Cov(X + a, Y + b) = \Cov(X, Y) $$

- The covariance of any random variable with a constant is $0$

  $$ \Cov(X, a) = 0 $$

- Covariances have the distributive property

  $$ \Cov(X + Z, Y) = \Cov(X, Y) + \Cov(Z, Y) $$

:::

## Covariance and correlation

::: {.incremental}

- Covariance is **scale dependent** - the units in which $X$ and $Y$ are measured matter for the overall magnitude of the covariance.
  - Sometimes we want a **scale free** measure of association as a summary.

- The **correlation** is constructed by normalizing the covariance between $X$ and $Y$ by their respective *standard deviations*

  $$\rho_{X,Y} = \Corr(X, Y) = \frac{\Cov(X, Y)}{\sqrt{\Var(X) \Var(Y)}} = \frac{\Cov(X, Y)}{\sqrt{\Var(X)} \sqrt{\Var(Y)}} $$

- The correlation coefficient lies between $[-1, 1]$
  - These bounds come from an application of the **Cauchy-Schwarz** inequality!

    $$\Cov(X, Y)^2 \le \Var(X) \Var(Y)$$

:::

## Correlation {.bvn-slide}

```{r}
#| echo: false

# A reprise of the opening animation's correlation steps: the same widget and
# script, opened straight onto the joint density at rho = 0, with the current
# rho in the slide title
p_rho <- htmlwidgets::onRender(
  bvn_base,
  paste(readLines("bvn-anim/bivariate_normal.js"), collapse = "\n"),
  data = list(grid = g, rhos = bvn_rhos, walk = lapply(bvn_walk, as.list),
              ztop2d = ztop_2d, ztop3d = ztop_3d,
              start = length(bvn_walk) + 1, title = "Correlation: ρ = {rho}")
)

# One empty fragment per correlation after the first
htmltools::tagList(
  p_rho,
  lapply(seq_len(length(bvn_rhos) - 1),
         function(i) htmltools::span(class = "fragment bvn-step"))
)
```

## Properties of variance

::: {.incremental}

- The **covariance** appears when we want to take the variance of a **sum** or **difference** in two random variables.

- The variance of the **sum** of two random variables is:

  $$ \Var(X + Y) = \Var(X) + \Var(Y) + 2\Cov(X, Y) $$

- The variance of the **difference** of two random variables is:

  $$ \Var(X - Y) = \Var(X) + \Var(Y) - 2\Cov(X, Y) $$

- **Intuitively**:
  - If I add two random variables together, I'm going to get something *noisier* unless $X$ tends to be high when $Y$ is low (and vice-versa)
  - If I subtract two random variables, I'm *also* going to get something *noisier* unless $X$ is high when $Y$ is high (and vice-versa)

:::

## Conditional distributions

::: {.incremental}

- We often want to ask how the distribution of one variable $Y$ **changes** if we were to set $X$ to a particular value
  - As we did before with conditional probabilities, we can define the **conditional distribution** of $Y \mid X$ 

- For **discrete** random variables, the definition is exactly what we had before: joint over marginal

  $$p_{Y \mid X}(y \mid x) = P(Y = y \mid X = x) = \frac{p_{XY}(x, y)}{p_X(x)} $$

- For **continuous** random variables it's a bit tricky (since $X = x$ has probability zero), but we use the same approach to define a conditional density

  $$f_{Y \mid X}(y \mid x) = \frac{f_{XY}(x, y)}{f_X(x)} $$

- Think of conditioning on $X = x$ as conditioning on a **tiny window** around $x$: that window appears in the numerator and denominator and cancels, leaving the ratio of densities.

:::

## Factoring joint PMFs/PDFs

::: {.incremental}

- As we did before with the probabilities of intersections of events, we can **factor** joint PMFs and PDFs into the product of conditionals

- For two discrete random variables $X$ and $Y$

  $$ p_{XY}(x, y) = p_{Y \mid X}(y \mid x) \cdot p_{X}(x) = p_{X \mid Y}(x \mid y) \cdot p_Y(y)$$

- For two continuous random variables $X$ and $Y$

  $$ f_{XY}(x, y) = f_{Y \mid X}(y \mid x) \cdot f_{X}(x) = f_{X \mid Y}(x \mid y) \cdot f_Y(y)$$

:::

## Factoring joint PMFs/PDFs

::: {.incremental}

- This generalizes to **many** random variables:

  - For discrete random variables $X_1, \dotsc, X_N$

    $$
    \begin{aligned}
    p_{X_1, \dotsc, X_N}(x_1, \dotsc, x_N) &= p_{X_1}(x_1) \cdot p_{X_2 \mid X_1}(x_2 \mid x_1) \cdot p_{X_3 \mid X_1, X_2}(x_3 \mid x_1, x_2) \\
    &\quad \cdots p_{X_N \mid X_1, \dotsc, X_{N-1}}(x_N \mid x_1, \dotsc, x_{N-1}) \\
    &= \prod_{i=1}^{N} p_{X_i \mid X_1, \dotsc, X_{i-1}}(x_i \mid x_1, \dotsc, x_{i-1})
    \end{aligned}
    $$

- For continuous random variables $X_1, \dotsc, X_N$

  $$
  \begin{aligned}
  f_{X_1, \dotsc, X_N}(x_1, \dotsc, x_N) &= f_{X_1}(x_1) \cdot f_{X_2 \mid X_1}(x_2 \mid x_1) \cdot f_{X_3 \mid X_1, X_2}(x_3 \mid x_1, x_2) \\
  &\quad \cdots f_{X_N \mid X_1, \dotsc, X_{N-1}}(x_N \mid x_1, \dotsc, x_{N-1}) \\
  &= \prod_{i=1}^{N} f_{X_i \mid X_1, \dotsc, X_{i-1}}(x_i \mid x_1, \dotsc, x_{i-1})
  \end{aligned}
  $$

:::

## Marginalizing joint distributions

::: {.incremental}

- We can recover a **marginal** distribution from a joint distribution by summing/integrating over whichever variable we want to get rid of
- To get the intuition, we can **factor** the joint density into a conditional and the marginal we care about

  $$f_Y(y) = \int_{-\infty}^{\infty} f_{XY}(x, y) dx = \int_{-\infty}^{\infty} f_{X \mid Y}(x \mid y) f_Y(y) dx $$
- Pull $f_Y(y)$ out of the integral (it doesn't depend on $x$)

  $$f_Y(y) = f_Y(y) \int_{-\infty}^{\infty} f_{X \mid Y}(x \mid y) dx $$

- And all densities integrate to $1$ (including conditional densities)

  $$f_Y(y) = f_Y(y) \times 1$$

:::


## Visualizing conditional distributions {.app-slide}

::: {.app-frame data-app="bivariate_normal_app.html" data-title="Slicing a bivariate normal"}
:::

```{=html}
<script>
// The widget is a separate page framed here. The <iframe> is built at runtime
// because `embed-resources: true` would otherwise inline the target into a
// data: URI, breaking the app's relative asset paths and its service worker.
// Mounting only once the slide is reached also delays the webR download.
(function () {
  function mount() {
    document.querySelectorAll("section.present .app-frame[data-app]").forEach(function (el) {
      if (el.dataset.mounted) return;
      el.dataset.mounted = "1";
      var f = document.createElement("iframe");
      f.src = el.dataset.app;
      f.title = el.dataset.title || "Interactive widget";
      f.setAttribute("allow", "clipboard-write");
      el.appendChild(f);
    });
  }
  // Reveal is not defined while this script is parsed, so wire up after load.
  window.addEventListener("load", function () {
    if (window.Reveal && typeof Reveal.on === "function") Reveal.on("slidechanged", mount);
    mount();
  });
})();
</script>
```

## Independence

::: {.incremental}

- Two random variables $X$ and $Y$ are **independent** if and only if their joint CDF factors into the product of their **marginal** CDFs
  - Essentially the same definition as before for events!

  $$ X \indep Y \iff F_{XY}(x, y) = F_X(x) \, F_Y(y) \quad \text{for all } x, y $$

- This also implies that their joint PMFs/PDFs factor as well

  $$ p_{XY}(x, y) = p_X(x) \, p_Y(y) \qquad f_{XY}(x, y) = f_X(x) \, f_Y(y) $$

- Conditioning on one doesn't change the distribution of the other 
  - The marginal and conditional distributions are the same

  $$ p_{Y \mid X}(y \mid x)  = p_Y(y) \qquad  p_{X \mid Y}(x \mid y) = p_{X}(x)$$
  $$ f_{Y \mid X}(y \mid x)  = f_Y(y) \qquad  f_{X \mid Y}(x \mid y) = f_{X}(x)$$

:::

## Independence and Covariance

::: {.incremental}

- Independence implies zero covariance

- First, let's write $\mathbb{E}[XY]$ as a double sum over the joint PMF using the law of the unconscious statistician.

  $$ \mathbb{E}[XY] = \sum_{x} \sum_{y} xy \cdot  p_{XY}(x, y) $$

- Independence means the joint distribution factors into the marginals

  $$ \mathbb{E}[XY] = \sum_{x} \sum_{y} xy \cdot  p_{X}(x) p_{Y}(y) $$

- $x p_X(x)$ factors out of our sum over $y$

  $$ \mathbb{E}[XY] = \sum_{x}x \cdot p_{X}(x) \sum_{y} y \cdot  p_{Y}(y) $$

:::

## Independence and Covariance

::: {.incremental}

- Plugging in our definitions of expected value

  $$
  \begin{aligned}
  \mathbb{E}[XY] &= \sum_{x} x \cdot p_{X}(x) \sum_{y} y \cdot p_{Y}(y) \\
  &= \sum_{x} x \cdot p_{X}(x) \, \mathbb{E}[Y] \\
  &= \mathbb{E}[Y] \sum_{x} x \cdot p_{X}(x) \\
  &= \mathbb{E}[X] \, \mathbb{E}[Y]
  \end{aligned}
  $$

- So the covariance is zero

  $$ \Cov(X, Y) = \mathbb{E}[XY] - \mathbb{E}[X] \, \mathbb{E}[Y] = \mathbb{E}[X] \, \mathbb{E}[Y] - \mathbb{E}[X] \, \mathbb{E}[Y] = 0 $$

:::


## Conditional expectations

::: {.incremental}

- We often want a **summary** of the conditional distribution of $Y$ given a particular value of $X$.
  - What is a good "prediction" for $Y$ if I know $X$?
- The **regression function** or **conditional expectation function** (CEF) is one such quantity

  $$
  \mu(x) = \mathbb{E}[Y \mid X = x] =
  \begin{cases}
  \displaystyle \sum_{y} y \cdot p_{Y \mid X}(y \mid x) & \text{discrete } Y \\[1em]
  \displaystyle \int_{-\infty}^{\infty} y \cdot f_{Y \mid X}(y \mid x) \, dy & \text{continuous } Y
  \end{cases}
  $$

- The CEF is a **function** of $X$
  - I give it a value of $X$, it outputs me what the expected value of $Y$ is for that value of $X$
- Learning about CEFs is at the heart of many social science questions!

:::

## How does Clinton's 2016 vote share vary by county? {.cef-slide}

```{r}
#| echo: false

# County-level 2016 presidential returns
election <- read.csv("data/prez_election_2016_vote.csv")
election$voteshare <- 100 * election$candidatevotes / election$totalvotes
election <- subset(election, !is.na(voteshare) & !is.na(pctWhiteNH))
cx <- election$pctWhiteNH
cy <- election$voteshare

# Joint density by a Gaussian kernel density estimate. Both variables live in
# [0, 100], and a plain KDE would leak mass past the edges (most counties sit
# just under 100% white), so every county is mirrored across 0 and 100 on each
# axis and the estimate is kept on [0, 100]^2 only.
mirror <- function(v) cbind(v, -v, 200 - v)
kx <- as.vector(mirror(cx)[, rep(1:3, 3)])
ky <- as.vector(mirror(cy)[, rep(1:3, each = 3)])
cg <- 0:100
f_xy <- MASS::kde2d(kx, ky, n = length(cg), lims = c(0, 100, 0, 100),
                    h = c(MASS::bandwidth.nrd(cx), MASS::bandwidth.nrd(cy)))$z
f_xy <- f_xy / sum(f_xy)   # f_xy[i, j] is the density at (x = cg[i], y = cg[j])

# Contours at the highest-density regions holding 10%, 20%, ..., 90% of the
# mass. Evenly spaced heights would all bunch up in the peak.
f_sorted <- sort(f_xy, decreasing = TRUE)
cef_levels <- sapply(seq(0.9, 0.1, -0.1),
                     function(p) f_sorted[which(cumsum(f_sorted) >= p)[1]])
cef_contours <- lapply(cef_levels, function(l) {
  segs <- contourLines(cg, cg, f_xy, levels = l)
  # one trace per level, separate rings split by NA gaps
  list(x = unlist(lapply(segs, function(s) c(s$x, NA))),
       y = unlist(lapply(segs, function(s) c(s$y, NA))))
})
level_cols <- colorRampPalette(c("#b89bd3", "#4a1766"))(length(cef_levels))

# The CEF of this joint density, E[Y | X = x] = sum_y y f(x, y) / sum_y f(x, y),
# and the BLP (the OLS line), each with the density's height along its path so
# it can ride on the surface. Drawn over the range of the data.
on_surface <- function(y_at) {
  keep <- cg >= floor(min(cx)) & cg <= ceiling(max(cx))
  z <- sapply(seq_along(cg), function(i) approx(cg, f_xy[i, ], y_at[i])$y)
  list(x = cg[keep], y = y_at[keep], z = z[keep])
}
cef_curve <- on_surface(as.vector(f_xy %*% cg) / rowSums(f_xy))
blp_fit <- coef(lm(voteshare ~ pctWhiteNH, data = election))
blp_curve <- on_surface(blp_fit[1] + blp_fit[2] * cg)

f_top <- max(f_xy)
cef_lift <- 0.012 * f_top   # keeps the lines just above the surface

# Slide title at each stage of the animation
# ("" keeps the slide's own heading)
cef_titles <- c("",
                "An empirical joint distribution",
                "An empirical joint distribution",
                "The conditional expectation function",
                "The conditional expectation function",
                "The CEF and the best linear predictor")

cef_axis <- function(title, ...) {
  list(title = list(text = title), gridcolor = "#d9d9d9", zeroline = FALSE,
       showspikes = FALSE, showbackground = TRUE, backgroundcolor = "#EFEFEF",
       tickfont = list(size = 12), tickangle = 0, ...)
}

cef_label <- function(x, y, text, colour) {
  list(x = x, y = y, z = cef_lift, text = text, showarrow = FALSE, xanchor = "left",
       font = list(size = 18, color = colour), visible = FALSE)
}

# Trace order matters: cef_joint.js addresses traces by index
p_cef <- plot_ly(width = 1000, height = 560) |>
  add_surface(x = cg, y = cg, z = t(f_xy), visible = FALSE,
              cmin = 0, cmax = f_top, showscale = FALSE, hoverinfo = "skip",
              # pale, so the contour lines and curves on it carry the picture
              colorscale = list(c(0, "#f4f0f8"), c(1, "#d6c4e6")),
              # mostly ambient light, so from overhead the surface stays pale
              # like a page rather than shading grey
              lighting = list(ambient = 1, diffuse = 0.12, specular = 0,
                              roughness = 1, fresnel = 0),
              contours = list(x = list(highlight = FALSE), y = list(highlight = FALSE),
                              z = list(highlight = FALSE))) |>
  add_trace(type = "scatter3d", mode = "markers", x = cx, y = cy, z = rep(0, length(cx)),
            marker = list(size = 2, color = "#333333"), hoverinfo = "skip")

for (j in seq_along(cef_contours)) {
  p_cef <- p_cef |>
    add_trace(type = "scatter3d", mode = "lines", x = cef_contours[[j]]$x,
              y = cef_contours[[j]]$y, z = rep(0, length(cef_contours[[j]]$x)),
              line = list(color = level_cols[j], width = 3), connectgaps = FALSE,
              hoverinfo = "skip", visible = FALSE)
}

p_cef <- p_cef |>
  add_trace(type = "scatter3d", mode = "lines", x = cef_curve$x, y = cef_curve$y,
            z = cef_curve$z, line = list(color = "#c5050c", width = 8),
            hoverinfo = "skip", visible = FALSE) |>
  add_trace(type = "scatter3d", mode = "lines", x = blp_curve$x, y = blp_curve$y,
            z = blp_curve$z, line = list(color = "#0479a8", width = 8),
            hoverinfo = "skip", visible = FALSE) |>
  layout(
    showlegend = FALSE,
    paper_bgcolor = "#F7F7F7",
    margin = list(l = 0, r = 0, t = 0, b = 0),
    font = list(family = "Red Hat Text, sans-serif", color = "#333333"),
    scene = list(
      xaxis = cef_axis("% non-Hispanic white", range = c(0, 100)),
      yaxis = cef_axis("Clinton vote share (%)", range = c(0, 100)),
      zaxis = cef_axis("", range = c(0, 1.05 * f_top), showticklabels = FALSE),
      aspectmode = "manual", aspectratio = list(x = 1, y = 1, z = 0.6),
      camera = list(projection = list(type = "orthographic")),   # set by the animation
      dragmode = "turntable", hovermode = FALSE,
      annotations = list(
        cef_label(3, 84, "CEF", "#c5050c"),
        cef_label(3, 58, "BLP", "#0479a8")
      )
    )
  ) |>
  config(displayModeBar = FALSE, doubleClick = FALSE)

p_cef <- htmlwidgets::onRender(
  p_cef,
  paste(readLines("cef-anim/cef_joint.js"), collapse = "\n"),
  data = list(grid = cg, density = t(f_xy), levels = cef_levels, lift = cef_lift,
              contours = cef_contours, cef = cef_curve, blp = blp_curve,
              points = list(x = cx, y = cy), titles = cef_titles, zaspect = 0.6)
)

# One empty fragment per animation step after the first
htmltools::tagList(
  p_cef,
  lapply(seq_len(length(cef_titles) - 1),
         function(i) htmltools::span(class = "fragment cef-step"))
)
```


## Law of Iterated Expectations

::: {.incremental}

- We can recover the marginal expectation $\mathbb{E}[Y]$ just by applying the expectation operator to $\mathbb{E}[Y \mid X]$
  - But what distribution are we taking that expectation over?
  - The distribution of $X$. Remember $\mathbb{E}[Y \mid X]$ is a function of $X$

- Applying the **law of the unconscious statistician**

  $$
  \mathbb{E}[Y] = \mathbb{E}\big[\mathbb{E}[Y \mid X]\big] =
  \begin{cases}
  \displaystyle \sum_{x} \mathbb{E}[Y \mid X = x] \cdot p_X(x) & \text{discrete } X \\[1em]
  \displaystyle \int_{-\infty}^{\infty} \mathbb{E}[Y \mid X = x] \cdot f_X(x) \, dx & \text{continuous } X
  \end{cases}
  $$

- This is known as the **law of total expectation** or **law of iterated expectation**

:::

## Law of Iterated Expectations

::: {.incremental}

- Let's complete the proof for the discrete case. Substitute in the definition of conditional expectation

  $$ \mathbb{E}\big[\mathbb{E}[Y \mid X]\big] =  \sum_{x} \sum_{y} y \cdot p_{Y \mid X}(y \mid x) \cdot p_X(x) $$

- That conditional times marginal is just the joint distribution of $X$ and $Y$ 

  $$ \mathbb{E}\big[\mathbb{E}[Y \mid X]\big] =  \sum_{x} \sum_{y} y \cdot p_{XY}(x,y)  $$

- Pull out the $y$ from the sum over $x$ (remember, we can swap order of summation)
  $$ \mathbb{E}\big[\mathbb{E}[Y \mid X]\big] =   \sum_{y} y \sum_{x} p_{XY}(x,y) $$

:::

## Law of Iterated Expectations

::: {.incremental}

- The sum over $x$ is just marginalizing out $x$ from the joint distribution of $X$ and $Y$
  $$
  \begin{aligned}
  \mathbb{E}\big[\mathbb{E}[Y \mid X]\big] &= \sum_{y} y \sum_{x} p_{XY}(x,y) \\
  &= \sum_{y} y \sum_{x} p_{X \mid Y}(x \mid y) \cdot p_Y(y) \\
  &= \sum_{y} y \cdot p_Y(y) \underbrace{\sum_{x} p_{X \mid Y}(x \mid y)}_{= 1} \\
  &= \sum_{y} y \cdot p_Y(y) = \mathbb{E}[Y]
  \end{aligned}
  $$

:::

## Law of Iterated Expectations

```{r}
#| echo: false
#| output: asis

# Inlined rather than linked so reveal.js sees the fragment groups inside
# (see assets/make_lie_slices.R): scatterplot, then slices, conditional means,
# shaded areas, the areas added, and the rearranged E[Y] rectangle
cat("```{=html}", readLines("assets/lie_slices.svg"), "```", sep = "\n")
```

## Example: Survey weighting to recalled vote

::: {.incremental}

- We want to predict the 2026 midterm elections.
  - **Problem**: Surveys aren't representative and we don't have a perfect way of dealing with **non-response bias**
- A newly popular strategy used by some polling firms is to **weight by recalled vote**
  - Imagine that the voters in 2026 are similar to 2024
  - We have a **known** 2024 benchmark: the observed Trump-Harris vote totals.
  - Calculate average vote choice **conditional** on recalled vote and average using the known 2024 distribution

:::

## Example: Weighting to recalled vote

::: {.incremental}

- **Recent example**: The Economist/YouGov poll (September 25–28, 2026) asks respondents 
  - $X$: the respondent's **recalled** 2024 vote: Trump or Harris
  - $Y_D$: 1 if they intend to vote for the **Democratic** House candidate in 2026, 0 otherwise
  - $Y_R$: 1 if they intend to vote for the **Republican** House candidate in 2026, 0 otherwise

- For each party's share:

  $$
  \begin{aligned}
  \mathbb{E}[Y_D] &= \mathbb{E}[Y_D \mid \text{Trump}] \, P(\text{Trump}) + \mathbb{E}[Y_D \mid \text{Harris}] \, P(\text{Harris}) \\
  \mathbb{E}[Y_R] &= \mathbb{E}[Y_R \mid \text{Trump}] \, P(\text{Trump}) + \mathbb{E}[Y_R \mid \text{Harris}] \, P(\text{Harris})
  \end{aligned}
  $$
:::

## Example: Weighting to recalled vote

```{r}
#| echo: false

# Economist/YouGov poll, September 25-28, 2026, registered voters, question 46
# ("In the elections for U.S. Congress in November, who will you vote for in
# the district where you live?"), crosstab by 2024 vote:
#   - dem, rep: shares choosing "The Democratic candidate" and "The Republican
#     candidate" (weighted, within each 2024 vote column)
#   - raw: unweighted respondents in each column (545 Trump, 586 Harris), as
#     shares of the two
#   - bench: YouGov's weighting target for 2024 presidential vote (48% Harris,
#     50% Trump, the actual result), rescaled to the two candidates
recall <- data.frame(
  x     = c("Trump", "Harris"),
  dem   = c(0.07, 0.90),
  rep   = c(0.80, 0.01),
  raw   = c(545, 586) / (545 + 586),
  bench = c(50, 48) / (50 + 48)
)
ey <- function(y, p) sum(recall[[y]] * recall[[p]])
f2 <- function(v) sprintf("%.2f", v)   # table cells, to two decimals
f3 <- function(v) sprintf("%.3f", v)
# The weighted average written out, e.g. 0.07 (0.51) + 0.90 (0.49)
terms <- function(y, p) paste0(f2(recall[[y]]), " (", f2(recall[[p]]), ")", collapse = " + ")
```

::: {.incremental}

- The Economist/YouGov crosstabs report:

  | 2024 vote ($x$) | $\mathbb{E}[Y_D \mid X = x]$ | $\mathbb{E}[Y_R \mid X = x]$ | $P(X = x)$, raw sample | $P(X = x)$, 2024 result |
  |:--|:-:|:-:|:-:|:-:|
  | `r recall$x[1]` | `r f2(recall$dem[1])` | `r f2(recall$rep[1])` | `r f2(recall$raw[1])` | `r f2(recall$bench[1])` |
  | `r recall$x[2]` | `r f2(recall$dem[2])` | `r f2(recall$rep[2])` | `r f2(recall$raw[2])` | `r f2(recall$bench[2])` |

- There are differences in how well each party is holding their voters in 2026
  - `r f2(recall$dem[2])` of Harris voters back the Democratic house candiate, but only `r f2(recall$rep[1])` of Trump voters back the Republican house candidate
  - Trump voters defect to the Democratic candidate (`r f2(recall$dem[1])`) far more than Harris voters defect to the Republican (`r f2(recall$rep[2])`)

- But the raw sample **over-represents** Harris voters (`r f2(recall$raw[2])` vs. `r f2(recall$bench[2])` in the actual result)

:::

## Example: Weighting to recalled vote

::: {.incremental}

- Averaging over the **2024 result**

  $$
  \begin{aligned}
  \mathbb{E}[Y_D] &= `r terms("dem", "bench")` = `r f3(ey("dem", "bench"))` \\
  \mathbb{E}[Y_R] &= `r terms("rep", "bench")` = `r f3(ey("rep", "bench"))`
  \end{aligned}
  $$

- The **margin**, $\mathbb{E}[Y_D] - \mathbb{E}[Y_R]$: the Democratic candidate leads by `r round(100 * (ey("dem", "bench") - ey("rep", "bench")))` points
  - The actual poll has it at $45\%$ to $37\%$ because they're also including 2024 non-voters who are leaning Democratic.

- **Survey weighting** is just the art of figuring out which distribution of $X$ to average over. 
  - In Week 13, we'll talk about how to actually do *estimation*

:::

## The mean minimizes mean squared error

::: {.incremental}

- Why do we focus so much on the **expected value** when thinking about prediction - why is it our "best guess"?
  - It turns out we can formalize the notion of "best guess" with respect to a particular error: the **mean squared error** (MSE)

- Let $c$ be our single guess. The mean squared error between $Y$ and $c$ is defined as $\mathbb{E}[(Y - c)^2]$
  - We want to find the value of $c$ that **minimizes** this quantity.

:::

## The mean minimizes mean squared error

::: {.incremental}

- Add and subtract $\mathbb{E}[Y]$ and expand the square

  $$
  \begin{aligned}
  \mathbb{E}[(Y - c)^2] &= \mathbb{E}\big[\big((Y - \mathbb{E}[Y]) + (\mathbb{E}[Y] - c)\big)^2\big] \\
  &= \mathbb{E}[(Y - \mathbb{E}[Y])^2] + 2(\mathbb{E}[Y] - c)\,\mathbb{E}[Y - \mathbb{E}[Y]] + (\mathbb{E}[Y] - c)^2 \\
  &= \Var(Y) + (\mathbb{E}[Y] - c)^2
  \end{aligned}
  $$

- The middle term drops out because $\mathbb{E}[Y - \mathbb{E}[Y]] = 0$

- $\Var(Y)$ doesn't depend on $c$, and $(\mathbb{E}[Y] - c)^2 \ge 0$, so the MSE is smallest at $c = \mathbb{E}[Y]$
  - The best constant prediction with respect to MSE is the **mean**, and the smallest achievable MSE is the **variance**

:::

## The CEF minimizes the MSE

::: {.incremental}

- Now, instead of a single constant $c$, we can plug in a **function** of $X$, $g(X)$
  - We can show that the function $g(X)$ that does the **best** job is the CEF. 
- The MSE between $Y$ and our function is $\mathbb{E}[(Y - g(X))^2]$ and we're searching over the space of all possible functions $g()$
- We'll use a similar trick from before: add and subtract the CEF $\mu(X) = \mathbb{E}[Y \mid X]$ and expand the square:

  $$
  \begin{aligned}
  \mathbb{E}[(Y - g(X))^2] &= \mathbb{E}[(Y - \mu(X))^2] + 2\,\mathbb{E}[(Y - \mu(X))(\mu(X) - g(X))] \\
  &\quad + \mathbb{E}[(\mu(X) - g(X))^2]
  \end{aligned}
  $$

:::

## The CEF minimizes the MSE

::: {.incremental}

- The middle term is zero: by the **law of total expectation**, condition on $X = x$ first. Given $X = x$, $\mu(x) - g(x)$ is just a constant

  $$ \mathbb{E}[(Y - \mu(x))(\mu(x) - g(x)) \mid X = x] = (\mu(x) - g(x))\big(\mathbb{E}[Y \mid X = x] - \mu(x)\big) = 0 $$

- So for any $g$

  $$ \mathbb{E}[(Y - g(X))^2] = \mathbb{E}[(Y - \mu(X))^2] + \underbrace{\mathbb{E}[(\mu(X) - g(X))^2]}_{\ge 0} \ge \mathbb{E}[(Y - \mu(X))^2] $$

- The CEF is the **best predictor** of $Y$ given $X$ in MSE.

:::

## Estimating the CEF: Clinton's 2016 vote share {.nostretch}

```{r}
#| echo: false

library(ggplot2)

# Binned scatterplot: 20 equal-count bins of % non-Hispanic white, with the
# mean of each variable within the bin (reuses `election` from the 3D slide)
election$bin <- cut(election$pctWhiteNH,
                    quantile(election$pctWhiteNH, seq(0, 1, length.out = 21)),
                    include.lowest = TRUE)
binned <- aggregate(cbind(pctWhiteNH, voteshare) ~ bin, data = election, FUN = mean)

cef_theme <- theme_minimal(base_size = 15) +
  theme(panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "#F7F7F7", colour = NA),
        panel.background = element_rect(fill = "#F7F7F7", colour = NA))

# The figure built up over the next few slides, one layer per step:
#   "counties"  the data, one point per county
#   "bins"      counties fade back; the bin means, an estimate of the CEF
#   "loess"     a second estimate of the same CEF, in the same red
#   "blp"       the BLP on top
# Each step's figure is a full plot on an opaque background, so stacking them
# as fragments (.r-stack) reads as one figure gaining layers.
cef_steps <- c("counties", "bins", "loess", "blp")
cef_plot <- function(step) {
  k <- match(step, cef_steps)
  p <- ggplot(election, aes(pctWhiteNH, voteshare)) +
    geom_point(size = 1, colour = "#333333", alpha = if (k == 1) 0.45 else 0.12)
  if (k >= 2) p <- p + geom_point(data = binned, size = 3, colour = "#c5050c")
  if (k >= 3) p <- p + geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
                                   colour = "#c5050c", linewidth = 1.4)
  if (k >= 4) p <- p + geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                                   colour = "#0479a8", linewidth = 1.4)
  p +
    scale_x_continuous(limits = c(0, 101), expand = c(0.01, 0)) +
    scale_y_continuous(limits = c(0, 101), expand = c(0.01, 0)) +
    labs(x = "% non-Hispanic white", y = "Clinton vote share (%)") +
    cef_theme
}
```

::: {.r-stack}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("counties")
```

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("bins")
```

:::

:::

## Estimating the CEF: Clinton's 2016 vote share {.nostretch}

::: {.r-stack}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("bins")
```

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("loess")
```

:::

:::

## How fast does the CEF change? {.nostretch}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center

# The same loess fit geom_smooth draws (span 0.75, degree 2), so its slopes can
# be read off: a centered difference at a handful of points
cef_fit <- loess(voteshare ~ pctWhiteNH, data = election)
at <- c(20, 40, 60, 80, 95)
mu_at <- function(x) predict(cef_fit, data.frame(pctWhiteNH = x))
slopes <- data.frame(x = at, y = mu_at(at), s = (mu_at(at + 1) - mu_at(at - 1)) / 2)
slopes$h <- pmin(9, 100 - at)   # half-width of each tangent, in points of % white;
                                # narrower near 100 so it stays on the axis
minus <- function(v, d = 2) sub("^-", "−", sprintf(paste0("%.", d, "f"), v))

# Tangents drawn heavier than the curve so they read as separate lines
ggplot(election, aes(pctWhiteNH, voteshare)) +
  geom_point(data = binned, size = 2.5, colour = "#c5050c", alpha = 0.35) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              colour = "#c5050c", linewidth = 0.9) +
  geom_segment(data = slopes, aes(x = x - h, xend = x + h,
                                  y = y - s * h, yend = y + s * h),
               colour = "#333333", linewidth = 2) +
  geom_point(data = slopes, aes(x, y), colour = "#333333", size = 3.5) +
  geom_text(data = slopes, aes(x, y + 11, label = minus(s)), size = 5.5,
            colour = "#333333") +
  scale_x_continuous(limits = c(0, 101), expand = c(0.01, 0)) +
  scale_y_continuous(limits = c(0, 101), expand = c(0.01, 0)) +
  labs(x = "% non-Hispanic white", y = "Clinton vote share (%)") +
  cef_theme
```


## The best linear predictor

::: {.incremental}

- The CEF can take **any** shape, which makes it hard to summarize when $X$ is continuous or high-dimensional.
  - Often we have hypotheses about a **rate of change** in an outcome as an independent variable increases or decreases
  - We want to estimate a **single number** to summarize an (average) rate of change.
- What if we **restrict** $g(X)$ to be a line, $g(X) = a + bX$?
- The **best linear predictor** (BLP) is the line with the smallest MSE

  $$ (\alpha, \beta) = \arg\min_{a,\, b} \ \mathbb{E}\big[(Y - a - bX)^2\big] $$

- We are searching over a **smaller** space of options

  $$ \mathbb{E}\big[(Y - \mu(X))^2\big] \le \mathbb{E}\big[(Y - \alpha - \beta X)^2\big] $$

:::

## The best linear predictor

::: {.incremental}

- Let's solve this optimization problem by taking the derivatives with respect to $a$ and $b$ and solving for $0$
- The derivative passes inside the expectation, then the chain rule: the outer square yields $2(Y - a - bX)$ and the inner function yields $-1$ for $a$ and $-X$ for $b$

  $$
  \begin{aligned}
  \frac{\partial}{\partial a} \mathbb{E}\big[(Y - a - bX)^2\big] &= \mathbb{E}\big[2(Y - a - bX) \cdot (-1)\big] = -2\,\mathbb{E}[Y - a - bX] \\
  \frac{\partial}{\partial b} \mathbb{E}\big[(Y - a - bX)^2\big] &= \mathbb{E}\big[2(Y - a - bX) \cdot (-X)\big] = -2\,\mathbb{E}\big[X(Y - a - bX)\big]
  \end{aligned}
  $$

- Set both to $0$ at $(a, b) = (\alpha, \beta)$ and divide both sides by $-2$

  $$ \mathbb{E}[Y - \alpha - \beta X] = 0 \qquad \mathbb{E}\big[X(Y - \alpha - \beta X)\big] = 0 $$

:::

## The best linear predictor

::: {.incremental}

- This gives the **first-order conditions** for the optimization problem

  $$ \mathbb{E}[Y - \alpha - \beta X] = 0 \qquad \mathbb{E}\big[X(Y - \alpha - \beta X)\big] = 0 $$

- The first gives the intercept: $\alpha = \mathbb{E}[Y] - \beta\,\mathbb{E}[X]$
- Plug it into the second

  $$
  \begin{aligned}
  0 &= \mathbb{E}[XY] - \alpha\,\mathbb{E}[X] - \beta\,\mathbb{E}[X^2] \\
  &= \mathbb{E}[XY] - \mathbb{E}[X]\,\mathbb{E}[Y] - \beta\big(\mathbb{E}[X^2] - \mathbb{E}[X]^2\big) \\
  &= \Cov(X, Y) - \beta \Var(X)
  \end{aligned}
  $$

- So the BLP is

  $$ \beta = \frac{\Cov(X, Y)}{\Var(X)} \qquad \alpha = \mathbb{E}[Y] - \beta\,\mathbb{E}[X] $$

:::

## The best linear predictor

::: {.incremental}

- Recall that for **any** function $g$

  $$ \mathbb{E}[(Y - g(X))^2] = \mathbb{E}[(Y - \mu(X))^2] + \mathbb{E}[(\mu(X) - g(X))^2] $$

- The first term doesn't depend on $g$. So among lines, minimizing the MSE for $Y$ is the same as minimizing the distance to the CEF

  $$ (\alpha, \beta) = \arg\min_{a,\, b} \ \mathbb{E}\big[(\mu(X) - a - bX)^2\big] $$

  - The BLP is the best **linear approximation to the CEF**
  - If the CEF is linear, the BLP **is** the CEF
- Like the CEF, the BLP is a feature of the joint distribution: an **estimand**
- For our Clinton vote share example: $\beta \approx `r sprintf("%.2f", coef(lm(voteshare ~ pctWhiteNH, data = election))[2])`$. 
  - Each additional increase in county % non-Hispanic white is associated with about half a point reduction in Clinton vote share

:::

## Estimating the BLP: Clinton's 2016 vote share {.nostretch}

::: {.r-stack}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("loess")
```

::: {.fragment}

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: center
cef_plot("blp")
```

:::

:::

## The BLP is a summary

```{r}
#| echo: false
#| output: asis

# Datasets that share a BLP, stepped through on each click: Anscombe's quartet,
# the Datasaurus, then made-up shapes on the dinosaur's means and BLP. Data
# from datasaurus-anim/make_datasaurus.R; drawn by datasaurus.js.
ds_cfg <- readLines("datasaurus-anim/datasaurus.json", warn = FALSE)
ds_js  <- readLines("datasaurus-anim/datasaurus.js")
n_ds   <- length(jsonlite::fromJSON(ds_cfg)$stages$name)
cat("```{=html}",
    '<div class="ds-anim"></div>',
    "<script>",
    "(", ds_js, ")(document.currentScript.previousElementSibling,", ds_cfg, ");",
    "</script>",
    # One empty fragment per animation step after the first
    rep('<span class="fragment ds-step"></span>', n_ds - 1),
    "```", sep = "\n")
```

::: {.attribution}
Anscombe's quartet from @anscombe1973graphs; the Datasaurus dozen from @matejka2017same
:::

## Bonus: The law of total variance

```{r}
#| echo: false

# The recalled vote example again: Y_D is binary, so each conditional variance
# is mu(x) (1 - mu(x)), averaged over the 2024 result
mu_d    <- recall$dem
p_x     <- recall$bench
ey_d    <- sum(mu_d * p_x)
within  <- sum(mu_d * (1 - mu_d) * p_x)   # E[Var(Y_D | X)]
between <- sum((mu_d - ey_d)^2 * p_x)     # Var(E[Y_D | X])
```

::: {.incremental}

- We can also decompose the variance of $Y$ by conditioning on $X$

  $$ \Var(Y) = \underbrace{\mathbb{E}\big[\Var(Y \mid X)\big]}_{\text{within}} + \underbrace{\Var\big(\mathbb{E}[Y \mid X]\big)}_{\text{between}} $$

- **Within**: the average spread of $Y$ around the CEF *inside* each slice of $X$
- **Between**: how much the CEF itself moves as $X$ changes. The part of the variation in $Y$ "explained" by $X$

- In the recalled vote example ($Y_D$ is binary, so $\Var(Y_D \mid X = x) = \mu(x)(1 - \mu(x))$)
  - Within: `r f3(within)`, between: `r f3(between)`, total: $\Var(Y_D) =$ `r f3(within + between)`
  - Recalled 2024 vote accounts for about `r round(100 * between / (within + between))`% of the variance in 2026 House vote preference

:::

## Bonus: The law of total variance

::: {.incremental}

- Start from $\Var(Y \mid X) = \mathbb{E}[Y^2 \mid X] - \mathbb{E}[Y \mid X]^2$ and take expectations of each piece, using the **law of iterated expectations**

  $$
  \begin{aligned}
  \mathbb{E}\big[\Var(Y \mid X)\big] &= \mathbb{E}\big[\mathbb{E}[Y^2 \mid X]\big] - \mathbb{E}\big[\mathbb{E}[Y \mid X]^2\big] = \mathbb{E}[Y^2] - \mathbb{E}\big[\mathbb{E}[Y \mid X]^2\big] \\
  \Var\big(\mathbb{E}[Y \mid X]\big) &= \mathbb{E}\big[\mathbb{E}[Y \mid X]^2\big] - \mathbb{E}\big[\mathbb{E}[Y \mid X]\big]^2 = \mathbb{E}\big[\mathbb{E}[Y \mid X]^2\big] - \mathbb{E}[Y]^2
  \end{aligned}
  $$

- Add the two lines: the $\mathbb{E}\big[\mathbb{E}[Y \mid X]^2\big]$ terms cancel

  $$ \mathbb{E}\big[\Var(Y \mid X)\big] + \Var\big(\mathbb{E}[Y \mid X]\big) = \mathbb{E}[Y^2] - \mathbb{E}[Y]^2 = \Var(Y) $$

- Since both pieces are non-negative, conditioning never *adds* variance: $\Var\big(\mathbb{E}[Y \mid X]\big) \le \Var(Y)$

:::

## Next week

::: {.incremental}

- What is your **estimand**?
  - How should we connect theory to empirical questions?
  - Thinking of estimands as features of distributions.
  - Preview of *descriptive* versus *causal* estimands
- **Midterm** exam
  - Probability questions in the *context* of actual research settings.
  - Focus on properties of conditional probability, properties of expectations. 

:::




## References {.refs-slide}

::: {#refs}
:::
